---
title: 'Exploring_Data.Rmd'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Reading the Data
We are interested in studying... so we will be making a plot for this.
```{r, include=FALSE}
setwd("/Users/pranavireddi/Documents/GitHub/SHHS")
data <- read.csv("first-shhs-dataset.csv")
dim(data)

library(ggplot2)

summary(data$gender)
data$gender <- as.factor(data$gender)
summary(data$gender)

levels(data$gender) <- c("male", "female")



neck_width <- data$NECK20
bmi <- data$bmi_s1
hip <- data$Hip
gender <- data$gender

age <- data$age_s1
oxygen_saturation <- data$avgsat
sleep_to_awake <- data$SlptAwP
time_in_bed <- data$TimeBedP
ankle_bp <- data$AnkBP
cholesterol <- data$Chol
triglycerides <- data$Trig
apnea_sleep_time <- data$pctSTApn
data$sleep_efficiency <- data$SlpEffP
atrial_fibrillation <- data$AFIB
height <- data$height

vars <- c("neck_width", "bmi", "hip", "gender", "age", "oxygen_saturation", "sleep_to_awake", "time_in_bed", "ankle_bp", "cholesterol", "triglycerides", "apnea_sleep_time", "sleep_efficiency", "atrial_fibrillation", "height")

```

## Histogram Plots

Below are the histogram and general plots for the bmi variable using ggplot.

```{r neck20, echo=FALSE}
ggplot(data, aes(x=bmi)) + 
  geom_histogram(nwidth=3, fill="#E69597", color="#E69597", alpha=0.9) +      
  theme(panel.background = element_rect(fill = "#B5D6D6", colour = "#B5D6D6", size   = 0.5, linetype = "solid")) +
      ggtitle("Histogram") +
      xlab("BMI") + ylab("Count")

```

## Overlaid Histogram Plots

Below are the overlaid histogram and general plots for the bmi variable and gender using ggplot.

```{r overlaid_plot, echo=FALSE}
ggplot(data, aes(x=bmi, color=gender, fill=gender)) + 
  geom_histogram(position="identity", alpha = 0.5) + 
  
    ggtitle("Overlaid Histogram") +
     xlab("BMI") + ylab("Count")

  theme(panel.background = element_rect(fill = "#B5D6D6", colour = "#B5D6D6", size        = 0.5, linetype = "solid")) 
```

## Boxplots

Below are the boxplots for the bmi variable and gender using ggplot.

```{r boxplot, echo=FALSE}
ggplot(data, aes(x=bmi, y=gender, color = gender, fill = gender)) + 
  geom_boxplot(position="identity") +
    ggtitle("Boxplot") +
      xlab("BMI") + ylab("Gender")

```

## Violin Plots

Below are the violin plots for the bmi variable and gender using ggplot.

```{r violin_plot, echo=FALSE}
ggplot(data, aes(x=bmi, y=gender, color = gender, fill = gender)) + 
  geom_violin(position="identity")+ 
    ggtitle("Violin Plot")+
      xlab("BMI")+ ylab("Gender")

```